### 2024-01-20 outputs/paper/paper.qmd
changes to all the script files

---
title: "In Depth Analysis of Event Patterns in Crisis Situations"
subtitle: "City of Toronto Data"
author: 
  - Vanshika Vanshika
thanks: "Code and data are available at: https://github.com/vanshikav2/Data_Analysis_1."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| message: false
#| echo: false

#### Workspace set-up ####

library(tidyverse)
library(palmerpenguins)
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(ggbeeswarm)
library(ggrepel)

#upload the cleaned_data
cleaned_data <- read_csv(
  file = here("outputs/data/analysis_data.csv"),
  show_col_types = FALSE)

#upload other datasets if needed

```

# Introduction

You can and should cross-reference sections and sub-sections.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

#ggplot(penguins, aes(x = island, fill = species)) +
#  geom_bar(alpha = 0.8) +
#  scale_fill_manual(values = c("darkorange","purple","cyan4"),
#                    guide = "none") +
#  theme_minimal() +
#  facet_wrap(~species, ncol = 1) +
#  coord_flip()
ggplot(cleaned_data, aes(x = event_type, fill = factor(hood_158_loc))) +
  geom_bar(position = "stack", width = 0.8) +
  labs(title = "Event Type Based on Hood_158", x = "Event Type", y = "Frequency", fill = "Hood_158")
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("outputs/models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-01-20 scripts/00-simulate_data.R

@@ -66,6 +66,6 @@ ggplot(simulated_data, aes(x = Event_type, fill = factor(Hood_158))) +
 
 # Visualize event types based on if the apprehension was made
 ggplot(simulated_data, aes(x = Event_type, fill = Apprehension_made)) +
-  geom_bar(position = "dodge") +
+  geom_bar(position = "dodge", fill = 'red') +
   labs(title = "Event Types Based on Apprehension Made", x = "Event Type", y = "Frequency", fill = "Apprehension Made")
 

### 2024-01-20 scripts/01-download_data.R

@@ -29,8 +29,7 @@ datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'
 # load the first datastore resource as a sample
 data <- filter(datastore_resources, row_number()==1) %>% get_resource()
 data
-
-
+dim(data)
 
 #### Save data ####
 # change the_raw_data to whatever name you assigned when you downloaded it.

### 2024-01-20 scripts/02-data_cleaning.R

@@ -12,32 +12,26 @@ library(tidyverse)
 #### Clean data ####
 raw_data <- read_csv("inputs/data/raw_data.csv")
 
+# cleans the data to include event_dow, type, if apprehension was made and their
+# hood_158 number
+
+raw_data
 cleaned_data <-
   raw_data |>
   janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
+  select(objectid, event_dow, event_type, apprehension_made, hood_158) |>
+  filter(hood_158 != "NSA") |>
+  rename(hood_158_loc = hood_158
          ) |> 
+  tidyr::drop_na()
+
+# create another cleaned dataset that contains all the dates and event type 
+cleaned_dated_data <-
+  raw_data |>
+  janitor::clean_names() |>
+  select(objectid, event_year, event_month, event_dow, event_type) |>
   tidyr::drop_na()
 
 #### Save data ####
 write_csv(cleaned_data, "outputs/data/analysis_data.csv")
+write_csv(cleaned_dated_data, "outputs/data/date_analysis_data.csv")

### 2024-01-20 scripts/03-test_data.R

@@ -1,18 +1,30 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests all cases of the simulation and data_cleaning.
+# Author: Vanshika Vanshika
+# Date: 19 January 2024
+# Contact: vanshika.vanshika@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: 00-simulate_data
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
-
+library(dplyr)
 #### Test data ####
 
 # get the data from 00-simulate_data to test
 source("scripts/00-simulate_data.R")
+source("scripts/02-data_cleaning.R")
+
+
+# check that there is max of 6 event types
+# reference from r bloggers: https://www.r-bloggers.com/2022/06/how-to-count-distinct-values-in-r/ 
+num_unique <- n_distinct(simulated_data$Event_type)
+ num_unique |> max() == 6
+
+# check the hood_158 is max 200 and min 100 
+simulated_data$Hood_158 |> min() == 100
+simulated_data$Hood_158 |> max() == 200
+
+# check in the cleaned data that there are 5 columns
+num_col = ncol(cleaned_data)
+num_col == 5

### 2024-01-21 outputs/paper/paper.qmd
Edits saturday

@@
 cleaned_data <- read_csv(
-  file = here("outputs/data/analysis_data.csv"),
+  file = here("outputs/data/analysis_data_full.csv"),
   show_col_types = FALSE)
 
-#upload other datasets if needed
+cleaned_data_hood <- read_csv(
+  file = here("outputs/data/analysis_data_hood158.csv"),
+  show_col_types = FALSE)
+
+cleaned_data_suicide <- read_csv(
+  file = here("outputs/data/suicide_data.csv"),
+  show_col_types = FALSE)
 
@@
 
-You can and should cross-reference sections and sub-sections.
+Toronto's Person In Crisis lines are an invaluable community asset that provides free, anonymous assistance to people experiencing a serious crisis.
+The frequency of crisis situations in today’s cities, particularly those involving populations at risk (PICs), highlights the need to better understand the dynamics behind such incidents.
+Over the years, the situation has drastically changed.
+As stated by Beverly Romeo-Beehler, Before the COVID-19 pandemic hit in 2018 and 2019, the call centre handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines.
+This is quite a large number of calls being received.
+He further adds that calls for service fell to 1.7 million in 2020 and then slightly increased to 1.8 million in 2021.
+During the pandemic in 2020 and 2021, slightly more than 60% of emergency 9-1-1 calls were placed.
+The call centre received over 5,000 calls a day on average between 2018 and 2021, with nearly 3,000 of those calls being 9-1-1 calls.
+
+The Toronto Police Service is in charge of the system, which is built to handle several kinds of crisis calls, each of which calls for a unique reaction.
+A variety of event types, including attempt suicide, person in crisis, elopee, jumper, overdose, and threatened suicide, are included in the dataset.The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014,2015.
+In this paper, all the neighborhood (Hood_158) data is used as an estimand to get a better understanding of all the hotspots of the PIC calls in Toronto neighborhood.
+In order to identify significant patterns and trends within the Persons in Crisis calls for help attended dataset, several examination of the interactions between type of events, geographical locations, month, day of the week & time of the day and event-specific aspect are used.
+The focus of this paper is to draw correlation between suicide related event and time of the day or day of the week.The analysis between suicide related events, their frequency and the geographical area of PIC call shows a definitive trend among these factors.
+The analysis aims to provide insightful information that will improve crisis response system and promote community well-being.
+
+In this data section, various aspects of dataset comparisons are analyzed, and the process of data cleaning used to derive those datasets is discussed.
+The results section of the paper presents all the results derived from different comparisons and datasets, and their analysis, trends, and understandings are discussed in the discussion section.
+Lastly, the conclusion section concludes the paper and provides final insights.
+Cross-referencing sections and sub-sections is implemented for coherence and clarity.
 
@@
 # Data {#sec-data}
+The data used in this paper is derived from the City of Toronto's opendatatoronto Portal().
+The dataset is titled 'Persons in Crisis Calls for Service Attended'.
+Data was cleaned and analyzed using the statistical programming software R [@r], and using other libraries such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `KableExtra` [@rKableExtra], `knitr` [@rknitr], `ggbeeswarm` [@rggbeeswarm], `ggrepel` [@rggrepel], and `here` [@rHere].
+Later in this paper, there will be a more thorough explanation of the procedures involved in collecting, cleaning, and analyzing data.
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+#Some of our data is of penguins (@fig-bills), from @palmerpenguins.
 
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| label: fig-events
+#| fig-cap: Types of Call Events
 #| echo: false
 
-#ggplot(penguins, aes(x = island, fill = species)) +
-#  geom_bar(alpha = 0.8) +
-#  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-#                    guide = "none") +
-#  theme_minimal() +
-#  facet_wrap(~species, ncol = 1) +
-#  coord_flip()
-ggplot(cleaned_data, aes(x = event_type, fill = factor(hood_158_loc))) +
-  geom_bar(position = "stack", width = 0.8) +
-  labs(title = "Event Type Based on Hood_158", x = "Event Type", y = "Frequency", fill = "Hood_158")
+# graph displays frequency of all the different type of events
+ ggplot(cleaned_data, aes(x = event_type)) +
+ geom_bar(fill = "pink") +
+ labs(title = "Event Types and Their Frequencies", x = "Event Type", y =      "Frequency", color = "Country") +
+ scale_color_brewer(palette = "Set1") +
+ theme(legend.position = "bottom")
+
+rows <- nrow(cleaned_data)
+rows
 ```
 
-Talk more about it.
+### Person In Crisis for Calls Attended
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+This dataset is collected and provided by the Toronto Police Service and the dataset was last updated January 11, 2024 as of January 21, 2024.
+It includes data from year 2014 to 2024.
+This dataset included event_id, the time,day,month, year of the call, the type of event that occured, if an apprehension after call was made, police division of event, whether an Occurrence was created or not.
+For the year 2014 and 2015, it also included the calls old and new name of neighbourhood, according to Hood_158 and Hood_140 system, which was left NSA after in order to protect the privacy of the parties.
+Upon collection of the data, it was cleaned in several different ways to analyse the data fully.
+The columns of neighbourhood addresses were removed, including division and event id.
+Please see @tbl-Full_data for the sample of this dataset
+```{r}
+#| message: false
+#| echo: false
+#| label: tbl-Full_data
+#| tbl-cap: Sample of Cleaned Person In Crisis for Calls Attended Data
+head(cleaned_data, 5) |>
+  kable(
+    col.names = c("Id","Event Year", "Event Month", "Event DOW", "Event Hour", "Event Type", "Apprehension made"),
+      booktabs = TRUE
+  )
+```
+### Hood_158 and Type of Event 
+Moreover, using the same data from Toronto Police Service, it was cleaned differently by deleting all the years after 2015 with no location, time,month,day of the evnt and adding hood_158 column of the data to show the geographical location of the calls in early years.
+A sample of this dataset can be viewed by @tbl-Hood
 
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| message: false
+#| echo: false
+#| label: tbl-Hood
+#| tbl-cap: Sample of Cleaned Person In Crisis With Hood_158 data 
+head(cleaned_data_hood, 5) |>
+  kable(
+    col.names = c("Id", "Event Type", "Hood_158 Number"),
+      booktabs = TRUE
+  )
+```
+
+```{r}
+#| label: fig-hood
+#| fig-cap: Relationship between suicide rate and the neighbourhood
 #| echo: false
@@
 #| message: false
+cleaned_data_hood <- read_csv(
+  file = here("outputs/data/analysis_data_hood158.csv"),
+  show_col_types = FALSE)
 
-analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))
+ggplot(cleaned_data_suicide, aes(x = factor(hood_158), fill = event_type)) +
+  geom_bar(position = "dodge", stat = "count") +
+  labs(title = "Event Type by Hood 158", x = "Hood 158", y = "Count") +
+  theme_minimal() +
+  theme(legend.position = "top",        # Change legend position
+        legend.title = element_blank()) +
+theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1, size = 10))
 
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
+category_freq <- table(cleaned_data_suicide$hood_158)
+
+# Find the value with the highest frequency (mode)
+sorted_values <- names(sort(category_freq, decreasing = TRUE))
+most_common_value <- sorted_values[1]
+second_most_common_value <- sorted_values[2]
+
+most_common_value
+second_most_common_value
+
+total <- sum(cleaned_data_suicide$hood_158 == "73")
+total
+unique_values_count <- length(unique(cleaned_data_suicide$hood_158))
+print(paste("Number of unique values in the 'Category' column:", unique_values_count))
 ```
@@
 
-# Model
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
 # Results
 
-Our results are summarized in @tbl-modelresults.
+### Type of Events ###
+From the years 2014 to 2023, Toronto has the highest number of PIC calls for the Person In crisis event which consists of all the different of type of crisis like elopee, jumper, and much more.
+The second highest category is suicide related which includes both suicide and suicide attempted and is a total of 91501 calls.
+The last category is Overdose.The total number of calls recorded in this dataset is 291991.
+The results can be seen @fig-events.
+
+### Neighbourhood and Suicide PIC Calls ###
+@fig-hood shows the relationship between neignbourhood and suicide PIC calls.
+There are around 174 neighbourhoods divided in Hood_158 system and the graph shows around 158 neighbourhoods have reported a PIC call for suicide-related events.
+The most number of calls were received from Hood 170 which were about 1022 suicide related calls and second most calls from recieved from the area Hood_73 which were around 858 calls.
+The Hood_170 is Yonge-Bay
+Corridor area and Hood_73 is Moss Park area in Old City of Toronto.(Toronto neigbourhoods #https://www.toronto.ca/city-government/data-research-maps/neighbourhoods-communities/neighbourhood-profiles/about-toronto-neighbourhoods/)
+
+### Suicides and Apprehension made ###
 
 ```{r}
+#| label: fig-planes2
 #| echo: false
@@
 library(rstanarm)
+library(patchwork)
+day_colors <- c("Sunday" = "red", "Monday" = "green", "Tuesday" = "blue", "Wednesday" = "purple",
+                "Thursday" = "orange", "Friday" = "brown", "Saturday" = "pink")
+# Create ggplot objects for days and time
+days <- cleaned_data %>%
+  ggplot(aes(x = fct_infreq(event_dow), fill = fct_infreq(event_dow))) +
+  geom_bar(position = "stack", show.legend = FALSE) +
+  labs(x = "Day of Week", y = "Frequency", fill = "Day of Week") +
+  scale_fill_manual(values = day_colors) +
+  theme_minimal()
 
-first_model <-
-  readRDS(file = here::here("outputs/models/first_model.rds"))
-```
+time <- cleaned_data %>%
+  mutate(Time=as.POSIXct(event_hour, format="%H")) %>%
+  ggplot(aes(Time, fill=fct_infreq(event_dow))) +
+  geom_density(adjust=0.5, position="stack") + # Stack the KDE of each day on top of each other
+  labs(x = "Time of Day", y = "Density", fill="Day of Week") +
+  scale_x_datetime(date_labels="%H", breaks="4 hour") +
+  scale_fill_manual(values=day_colors) +
+  theme(legend.margin=margin(0,0,0,0),
+        legend.box.margin=margin(-5,-5,-5,-5),
+        axis.text.x=element_text(size=8)) +
+  ggtitle("Delays Spike Several Times Throughout the Day")
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
+# Plot the two graphs on top of each other with a space in between
+(days / plot_spacer() / time +
+    plot_layout(heights = c(10, 1, 6), guides="collect"))
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
 ```
@@
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
+#pp_check(first_model) +
+#  theme_classic() +
+#  theme(legend.position = "bottom")
 
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
+#posterior_vs_prior(first_model) +
+#  theme_minimal() +
+ # scale_color_brewer(palette = "Set1") +
+ # theme(legend.position = "bottom") +
+ # coord_flip()
 ```
@@
 
-plot(first_model, "trace")
+#plot(first_model, "trace")
 
-plot(first_model, "rhat")
+#plot(first_model, "rhat")
 ```

### 2024-01-21 scripts/00-simulate_data.R

@@ -49,6 +49,7 @@ ggplot(simulated_data, aes(x = Event_type)) +
 scale_color_brewer(palette = "Set1") +
   theme(legend.position = "bottom")
 
+
 # Visualize event types based on neighborhoods
 #ggplot(simulated_data, aes(x = Event_type, fill = NEIGHBOURHOOD)) +
 #  geom_bar(position = "dodge") +

### 2024-01-21 scripts/01-download_data.R

@@ -19,15 +19,15 @@ library(readr)
 # get package
 package <- show_package("persons-in-crisis-calls-for-service-attended")
 package
-
+dim(package)
 # get all resources for this package
 resources <- list_package_resources("persons-in-crisis-calls-for-service-attended")
-
+resources
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
 datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
 
 # load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+data <- filter(datastore_resources, row_number()==2) %>% get_resource()
 data
 dim(data)
 

### 2024-01-21 scripts/02-data_cleaning.R

@@ -12,26 +12,53 @@ library(tidyverse)
 #### Clean data ####
 raw_data <- read_csv("inputs/data/raw_data.csv")
 
-# cleans the data to include event_dow, type, if apprehension was made and their
-# hood_158 number
-
+# cleans the data to include event type and hood_158 number
 raw_data
-cleaned_data <-
+cleaned_data_1<-
   raw_data |>
   janitor::clean_names() |>
-  select(objectid, event_dow, event_type, apprehension_made, hood_158) |>
+  select(objectid, event_type, hood_158) |>
   filter(hood_158 != "NSA") |>
   rename(hood_158_loc = hood_158
          ) |> 
   tidyr::drop_na()
 
-# create another cleaned dataset that contains all the dates and event type 
-cleaned_dated_data <-
+# create cleaned dataset that contains all the dates from year 2014 
+cleaned_data_2 <-
   raw_data |>
   janitor::clean_names() |>
   select(objectid, event_year, event_month, event_dow, event_type) |>
+  filter(event_year == "2014") |>
   tidyr::drop_na()
 
+# create another cleaned dataset that contains all the dates from year 2023
+cleaned_data_3 <-
+  raw_data |>
+  janitor::clean_names() |>
+  select(objectid, event_year, event_month, event_dow, event_type) |>
+  filter(event_year == "2023") |>
+  tidyr::drop_na()
+
+# create cleaned dataset that contains all the dates and event type 
+cleaned_data_4 <-
+  raw_data |>
+  janitor::clean_names() |>
+  select(objectid, event_year, event_month, event_dow,event_hour, event_type, 
+         apprehension_made) |>
+  tidyr::drop_na()
+
+# create cleaned dataset that contains all the hood_158_loc of Suicides
+event_suicide_data <-
+  raw_data |>
+  janitor::clean_names() |>
+  select(objectid, event_type, hood_158) |>
+  filter(event_type == "Suicide-related" & hood_158 != "NSA") |>
+  tidyr::drop_na()
+
+
 #### Save data ####
-write_csv(cleaned_data, "outputs/data/analysis_data.csv")
-write_csv(cleaned_dated_data, "outputs/data/date_analysis_data.csv")
+write_csv(cleaned_data_1, "outputs/data/analysis_data_hood158.csv")
+write_csv(cleaned_data_2, "outputs/data/analysis_data_2014.csv")
[…]

### 2024-01-22 outputs/paper/paper.qmd
work sunday

@@
 ### Neighbourhood and Suicide PIC Calls ###
-@fig-hood shows the relationship between neignbourhood and suicide PIC calls.
-There are around 174 neighbourhoods divided in Hood_158 system and the graph shows around 158 neighbourhoods have reported a PIC call for suicide-related events.
-The most number of calls were received from Hood 170 which were about 1022 suicide related calls and second most calls from recieved from the area Hood_73 which were around 858 calls.
+@fig-hood shows the relationship between neighborhood and suicide PIC calls.
+There are around 174 neighborhoods divided in Hood_158 system and the graph shows around 158 neighborhoods have reported a PIC call for suicide-related events.
+The most number of calls were received from Hood 170 which were about 1022 suicide related calls and second most calls from received from the area Hood_73 which were around 858 calls.
 The Hood_170 is Yonge-Bay
-Corridor area and Hood_73 is Moss Park area in Old City of Toronto.(Toronto neigbourhoods #https://www.toronto.ca/city-government/data-research-maps/neighbourhoods-communities/neighbourhood-profiles/about-toronto-neighbourhoods/)
+Corridor area and Hood_73 is Moss Park area in Old City of Toronto.(Toronto neighborhoods #https://www.toronto.ca/city-government/data-research-maps/neighbourhoods-communities/neighbourhood-profiles/about-toronto-neighbourhoods/)
 
-### Suicides and Apprehension made ###
-
+### Suicide calls for different days and time of the day ###
+The graph shows the time of the day that was most frequent for the suicide calls and day of the week that most people called for suicide PIC help. 
 ```{r}
@@
 # Create ggplot objects for days and time
-days <- cleaned_data %>%
-  ggplot(aes(x = fct_infreq(event_dow), fill = fct_infreq(event_dow))) +
-  geom_bar(position = "stack", show.legend = FALSE) +
-  labs(x = "Day of Week", y = "Frequency", fill = "Day of Week") +
-  scale_fill_manual(values = day_colors) +
-  theme_minimal()
 
@@
   ggplot(aes(Time, fill=fct_infreq(event_dow))) +
-  geom_density(adjust=0.5, position="stack") + # Stack the KDE of each day on top of each other
-  labs(x = "Time of Day", y = "Density", fill="Day of Week") +
-  scale_x_datetime(date_labels="%H", breaks="4 hour") +
+  geom_density(adjust=0.5, position="stack") +
+  labs(x = "Hour", y = "Density", fill= "Day of Week") +
+  scale_x_continuous(breaks = seq(0, 24, by = 2)) +
   scale_fill_manual(values=day_colors) +
-  theme(legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-5,-5,-5,-5),
-        axis.text.x=element_text(size=8)) +
-  ggtitle("Delays Spike Several Times Throughout the Day")
+  theme(axis.text.x=element_text(size=8)) +
+  ggtitle("Frequency of Suicide Calls")
 
-# Plot the two graphs on top of each other with a space in between
-(days / plot_spacer() / time +
-    plot_layout(heights = c(10, 1, 6), guides="collect"))
+time
 

### 2024-01-22 outputs/paper/paper.qmd
-Cleaned and made another dataset -Almost finished the Quarto document -finished all the references

@@
 library(tidyverse)
-library(palmerpenguins)
 library(knitr)
@@
 
+cleaned_data_suicide1 <- read_csv(
+  file = here("outputs/data/suicide_data1.csv"),
+  show_col_types = FALSE)
+
 ```
@@
 Over the years, the situation has drastically changed.
-As stated by Beverly Romeo-Beehler, Before the COVID-19 pandemic hit in 2018 and 2019, the call centre handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines.
+As stated by Beverly Romeo-Beehler [@resource1], Before the COVID-19 pandemic hit in 2018 and 2019, the call centre handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines.
 This is quite a large number of calls being received.
@@
 # Data {#sec-data}
-The data used in this paper is derived from the City of Toronto's opendatatoronto Portal().
+The data used in this paper is derived from the City of Toronto's opendatatoronto Portal [@torontodata].
 The dataset is titled 'Persons in Crisis Calls for Service Attended'.
-Data was cleaned and analyzed using the statistical programming software R [@r], and using other libraries such as `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `KableExtra` [@rKableExtra], `knitr` [@rknitr], `ggbeeswarm` [@rggbeeswarm], `ggrepel` [@rggrepel], and `here` [@rHere].
+Data was cleaned and analyzed using the statistical programming software R [@citeR], and using other libraries such as `tidyverse` [@tidy], `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan],`reshape2`` [@reshape], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], and `here` [@here].
 Later in this paper, there will be a more thorough explanation of the procedures involved in collecting, cleaning, and analyzing data.
-
-#Some of our data is of penguins (@fig-bills), from @palmerpenguins.
 
@@
  theme(legend.position = "bottom")
-
-rows <- nrow(cleaned_data)
-rows
 ```
@@
 #| tbl-cap: Sample of Cleaned Person In Crisis for Calls Attended Data
+
+# creates a table of the cleaned sample data set
 head(cleaned_data, 5) |>
@@
 ### Hood_158 and Type of Event 
-Moreover, using the same data from Toronto Police Service, it was cleaned differently by deleting all the years after 2015 with no location, time,month,day of the evnt and adding hood_158 column of the data to show the geographical location of the calls in early years.
-A sample of this dataset can be viewed by @tbl-Hood
+Moreover, using the same data from Toronto Police Service, it was cleaned differently by deleting all the years after 2015 with no location, time,month,day of the event and adding hood_158 column of the data to show the geographical location of the calls in early years.
+A sample of this data set can be viewed by @tbl-Hood
 
@@
 #| tbl-cap: Sample of Cleaned Person In Crisis With Hood_158 data
+
+# creates a data set of the cleaned data set of the neighbourhoods
 head(cleaned_data_hood, 5) |>
@@
 #| message: false
+
+# reads a different csv file
 cleaned_data_hood <- read_csv(
@@
 
+# makes a plot of all the frequency of the suicide related calls in each neighbourhood
 ggplot(cleaned_data_suicide, aes(x = factor(hood_158), fill = event_type)) +
@@
         legend.title = element_blank()) +
-theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1, size = 10))
-
-category_freq <- table(cleaned_data_suicide$hood_158)
-
-# Find the value with the highest frequency (mode)
-sorted_values <- names(sort(category_freq, decreasing = TRUE))
-most_common_value <- sorted_values[1]
-second_most_common_value <- sorted_values[2]
-
-most_common_value
-second_most_common_value
-
-total <- sum(cleaned_data_suicide$hood_158 == "73")
-total
-unique_values_count <- length(unique(cleaned_data_suicide$hood_158))
-print(paste("Number of unique values in the 'Category' column:", unique_values_count))
+theme(axis.text.x=element_text(size=0)) # to not show the x-axis as too many numbers
 ```
-
-Talk way more about it.
 
@@
 The Hood_170 is Yonge-Bay
-Corridor area and Hood_73 is Moss Park area in Old City of Toronto.(Toronto neighborhoods #https://www.toronto.ca/city-government/data-research-maps/neighbourhoods-communities/neighbourhood-profiles/about-toronto-neighbourhoods/)
+Corridor area and Hood_73 is Moss Park area in Old City of Toronto[@resource2].
 
 ### Suicide calls for different days and time of the day ###
-The graph shows the time of the day that was most frequent for the suicide calls and day of the week that most people called for suicide PIC help. 
+The graph shows the time of the day that was most frequent for the suicide calls and day of the week that most people called for suicide PIC help.This is shown in [@Fig-days_and_time]
 ```{r}
-#| label: fig-planes2
+#| label: fig-days_and_time
 #| echo: false
@@
 #| message: false
+#| fig-cap: Frequency of Calls for Day and Time
 
-library(rstanarm)
-library(patchwork)
 day_colors <- c("Sunday" = "red", "Monday" = "green", "Tuesday" = "blue", "Wednesday" = "purple",
@@
 
-## First discussion point {#sec-first-point}
+Upon thorough examination of the dataset, several significant findings emerge.
+Overall, suicide PIC calls show a very important and cruicial relationship with the other factors such as time of the day, day of the week, and geographical area.
+As highlighted in the Canadian Journal of Psychiatry[@resource3], suicide remains a noteworthy cause of death in Toronto, necessitating a nuanced understanding of the factors contributing to suicide and the development of targeted prevention strategies.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
-
-## Second discussion point
-
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
-
-\newpage
-
-\appendix
-
-# Appendix {-}
-
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+The analysis of PIC calls reveals a persistent concern, as suicide-related calls consistently constitute a substantial portion of all recorded PIC calls over the years.
+This trend, observed in both 2014 and 2023, emphasizes the ongoing importance of addressing mental health crises within the community.
+Mapping these calls to specific areas reveals distinct hotspots, with neighborhoods such as Yonge-Bay Corridor (Hood_170) and Moss Park in the Old City of Toronto (Hood_73) reporting a disproportionately high number of suicide-related PIC calls, indicating potential areas for targeted intervention and support.
 
 ```{r}
+#| label: fig-correlation
+#| echo: false
 #| eval: true
-#| echo: false
+#| warning: false
 #| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+#| fig-cap: Correlation of Suicide with Time of the Day
 
-#pp_check(first_model) +
-#  theme_classic() +
-#  theme(legend.position = "bottom")
+library(reshape2)
 
-#posterior_vs_prior(first_model) +
-#  theme_minimal() +
- # scale_color_brewer(palette = "Set1") +
- # theme(legend.position = "bottom") +
- # coord_flip()
+# creates a correlation graph between time of the day and Suicide calls
+# code inspiration from https://stackoverflow.com/questions/66652330/error-ggplot-continuous-x-aesthetic-did-you-forget-aesgroup
+cleaned_data_suicide1 |>
+    filter(event_type == "Suicide-related") %>%
+  group_by(event_hour) %>% # calculating by hours
+  summarise(suicide_frequency = n()) %>% # to get the frequency of the suicide calls
+  ggplot(aes(x = factor(event_hour), y = suicide_frequency, label = "Calls")) + geom_point() + 
+  ggrepel::geom_text_repel() + 
+labs(
+    title = "Frequency of PIC Calls By The Hour",
+    x = "Hour of the Day",
+    y = "Frequency of Calls"
+  ) +
+  theme_classic()
+
+table_data <- table(cleaned_data_suicide1$event_dow, cleaned_data_suicide1$event_hour)
+
+# Convert the table to a data frame
+#df <- as.data.frame.matrix(table_data)
+
+# Reorder days of the week
+#df$Day <- factor(row.names(df), levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+
+# Melt the data for ggplot
+#melted_df <- melt(df, id.vars = "Day")
+
+# Create a heatmap
+#ggplot(melted_df, aes(x = variable, y = Day, fill = value)) +
+ # geom_tile() +
+#  scale_fill_gradient(low = "white", high = "darkred") +
+ # labs(title = "Suicide Calls by Day of the Week and Hour of the Day",
+#       x = "Hour of the Day",
+ #      y = "Day of the Week") +
+ # theme_minimal()
+
 ```
 
-## Diagnostics
+There is also a clear a correlation between the time of the day and day of the week with the frequency of suicide-related PIC calls like [@fig-correlation].
+Thursday is the day of the week that receives most of the calls with Sunday being the day that receives the least calls.
+The hours between 1300 to 1500 receive a high amount of calls compared to the other time of the day which shows a direct relation between the day and time and suicide related calls.
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+However, it is crucial to acknowledge and discuss the inherent biases and limitations within the dataset.
+The presence of numerous "NAS" (Not A Statement) assigned rows in the raw data introduces the possibility of errors in the graphs and analyses, potentially diverting the accuracy of the findings.
+Additionally, the dataset lacks explicit information regarding the reasons for caller apprehension, hindering a direct correlation between apprehension and the nature of events.
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+Moreover, the 10-year span covered by the dataset introduces potential biases due to changes in data collection techniques and technology over time.
+The evolving landscape of mental health awareness and reporting practices may influence the comparability of older and more recent data.
+Furthermore, limitations arise from the cessation of recording neighborhood information after 2015, aimed at safeguarding the privacy of affected individuals.
+This restriction poses challenges in assessing the situation in 2023 and analyzing whether observed patterns persist or have evolved.
 
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
+## Conclusion ##
 
-#plot(first_model, "trace")
+Finally, the examination of Person In Crisis (PIC) calls in Toronto offers insightful information about the dynamics of mental health crises in the community.
+The dataset, which covers the years 2014 to 2024, shows a steady and noteworthy increase in PIC calls connected to suicide, highlighting the continued need for focused intervention and support.
+The necessity of nuanced techniques in crisis management is shown by the association between suicide-related incidents and the time of day, day of the week, and geographic location.
 
-#plot(first_model, "rhat")
-```
-
+Overall, this analysis provides insightful data that will improve Toronto's crisis response system's efficacy.
+The results underline how crucial it is to continue treating mental health emergencies and lay the groundwork for next initiatives to raise community well-being.
 \newpage

### 2024-01-22 scripts/02-data_cleaning.R

@@ -55,6 +55,13 @@ event_suicide_data <-
   filter(event_type == "Suicide-related" & hood_158 != "NSA") |>
   tidyr::drop_na()
 
+event_suicide_data2 <-
+raw_data |>
+  janitor::clean_names() |>
+  select(objectid, event_month, event_dow, event_hour, event_type) |>
+  filter(event_type == "Suicide-related") |>
+  tidyr::drop_na()
+print(event_suicide_data2)
 
 #### Save data ####
 write_csv(cleaned_data_1, "outputs/data/analysis_data_hood158.csv")
@@ -62,3 +69,5 @@ write_csv(cleaned_data_2, "outputs/data/analysis_data_2014.csv")
 write_csv(cleaned_data_3, "outputs/data/analysis_data_2023.csv")
 write_csv(cleaned_data_4, "outputs/data/analysis_data_full.csv")
 write_csv(event_suicide_data, "outputs/data/suicide_data.csv")
+write_csv(event_suicide_data2, "outputs/data/suicide_data1.csv")
+

### 2024-01-22 outputs/paper/paper.qmd
- Edited the README page to make it look more professional - Added the Abstract on the paper.

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Toronto Police Services' 10-year-long dataset on Persons in Crisis Calls for Service Attended was used to conduct an in-depth analysis of the major PIC call event: Suicide and its data characteristics.
+It is found that there is a direct relation/trend that can be observed between suicide PIC calls and several factors such as the specific time of the day, day of the week, and identifed hotspot geographical location.
+This is important as it's essential to analyze the patterns of suicide PIC calls to effectively deal with the problem at its root."
 format: pdf

### 2024-01-22 outputs/paper/paper.qmd
Some minor error fixes

@@
 The dataset is titled 'Persons in Crisis Calls for Service Attended'.
-Data was cleaned and analyzed using the statistical programming software R [@citeR], and using other libraries such as `tidyverse` [@tidy], `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan],`reshape2`` [@reshape], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], and `here` [@here].
+Data was cleaned and analyzed using the statistical programming software R [@citeR], and using other libraries such as `tidyverse` [@tidy], `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan],`reshape2` [@reshape], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], and `here` [@here].
 Later in this paper, there will be a more thorough explanation of the procedures involved in collecting, cleaning, and analyzing data.

### 2024-01-23 outputs/paper/paper.qmd
Final changes to the file. Updated README. Entered the LLM usage. Final paper and pdf.

@@
 ---
-title: "In Depth Analysis of Event Patterns in Crisis Situations"
+title: "Analysing Suicide Event Patterns Show Trend amongst Crisis Calls"
 subtitle: "City of Toronto Data"
@@
 format: pdf
+toc: true
 number-sections: true
@@
 Toronto's Person In Crisis lines are an invaluable community asset that provides free, anonymous assistance to people experiencing a serious crisis.
-The frequency of crisis situations in today’s cities, particularly those involving populations at risk (PICs), highlights the need to better understand the dynamics behind such incidents.
+The frequency of crises in today’s cities, particularly those involving populations at risk (PICs), highlights the need to better understand the dynamics behind such incidents.
 Over the years, the situation has drastically changed.
-As stated by Beverly Romeo-Beehler [@resource1], Before the COVID-19 pandemic hit in 2018 and 2019, the call centre handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines.
+As stated by Beverly Romeo-Beehler [@resource1], Before the COVID-19 pandemic hit in 2018 and 2019, the call center handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines.
 This is quite a large number of calls being received.
 He further adds that calls for service fell to 1.7 million in 2020 and then slightly increased to 1.8 million in 2021.
-During the pandemic in 2020 and 2021, slightly more than 60% of emergency 9-1-1 calls were placed.
-The call centre received over 5,000 calls a day on average between 2018 and 2021, with nearly 3,000 of those calls being 9-1-1 calls.
 
 The Toronto Police Service is in charge of the system, which is built to handle several kinds of crisis calls, each of which calls for a unique reaction.
-A variety of event types, including attempt suicide, person in crisis, elopee, jumper, overdose, and threatened suicide, are included in the dataset.The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014,2015.
-In this paper, all the neighborhood (Hood_158) data is used as an estimand to get a better understanding of all the hotspots of the PIC calls in Toronto neighborhood.
-In order to identify significant patterns and trends within the Persons in Crisis calls for help attended dataset, several examination of the interactions between type of events, geographical locations, month, day of the week & time of the day and event-specific aspect are used.
-The focus of this paper is to draw correlation between suicide related event and time of the day or day of the week.The analysis between suicide related events, their frequency and the geographical area of PIC call shows a definitive trend among these factors.
-The analysis aims to provide insightful information that will improve crisis response system and promote community well-being.
+A variety of event types, including attempt suicide, the person in crisis, elopee, jumper, overdose, and threatened suicide, are included in the data set.
+The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014,2015.
+In this paper, all the neighborhood (Hood_158) data is used as an estimand to get a better understanding of all the hotspots of the PIC calls in the Toronto neighborhood.
+To identify significant patterns and trends within the Persons in Crisis calls for help attended data set, several examinations of the interactions between type of events, geographical locations, month, day of the week & time of the day, and event-specific aspects are used.
+The focus of this paper is to draw the correlation between suicide-related events and the time of the day or day of the week.
+The analysis of suicide-related events, their frequency, and the geographical area of PIC calls shows a definitive trend among these factors.
+The analysis aims to provide insightful information that will improve the crisis response system and promote community well-being.
 
-In this data section, various aspects of dataset comparisons are analyzed, and the process of data cleaning used to derive those datasets is discussed.
-The results section of the paper presents all the results derived from different comparisons and datasets, and their analysis, trends, and understandings are discussed in the discussion section.
+In this data section, various aspects of data set comparisons are analyzed, and the data-cleaning process used to derive those data sets is discussed.
+The results section of the paper presents all the results derived from different comparisons and data sets, and their analysis, trends, and understandings are discussed in the discussion section.
 Lastly, the conclusion section concludes the paper and provides final insights.
-Cross-referencing sections and sub-sections is implemented for coherence and clarity.
+Cross-referencing sections and sub-sections are implemented for coherence and clarity.
 
@@
 
-This dataset is collected and provided by the Toronto Police Service and the dataset was last updated January 11, 2024 as of January 21, 2024.
-It includes data from year 2014 to 2024.
-This dataset included event_id, the time,day,month, year of the call, the type of event that occured, if an apprehension after call was made, police division of event, whether an Occurrence was created or not.
-For the year 2014 and 2015, it also included the calls old and new name of neighbourhood, according to Hood_158 and Hood_140 system, which was left NSA after in order to protect the privacy of the parties.
-Upon collection of the data, it was cleaned in several different ways to analyse the data fully.
-The columns of neighbourhood addresses were removed, including division and event id.
+This dataset is collected and provided by the Toronto Police Service and the dataset was last updated January 11, 2024, as of January 21, 2024.
+It includes data from the year 2014 to 2024.
+This dataset included event_id, the time, day, month, and year of the call, the type of event that occurred if an apprehension after the call was made, the police division of the event, and whether an Occurrence was created.
+For the years 2014 and 2015, it also included the calls for old and new names of neighborhood, according to the Hood_158 and Hood_140 system, which was left “NSA” after to protect the privacy of the parties.
+Upon collection of the data, it was cleaned in several different ways to analyze the data fully.
+The columns of neighborhood addresses were removed, including division and event id.
 Please see @tbl-Full_data for the sample of this dataset
@@
 ### Hood_158 and Type of Event 
-Moreover, using the same data from Toronto Police Service, it was cleaned differently by deleting all the years after 2015 with no location, time,month,day of the event and adding hood_158 column of the data to show the geographical location of the calls in early years.
-A sample of this data set can be viewed by @tbl-Hood
+Moreover, using the same data from the Toronto Police Service, it was cleaned differently by deleting all the years after 2015 with no location,  time, month, or day of the event and adding the hood_158 column of the data to show the geographical location of the calls in early years.
+A sample of this data set can be viewed by @tbl-Hood.
 
@@
 
-# creates a data set of the cleaned data set of the neighbourhoods
+# creates a data set of the cleaned data set of the neighborhoods
 head(cleaned_data_hood, 5) |>
@@
 ### Type of Events ###
-From the years 2014 to 2023, Toronto has the highest number of PIC calls for the Person In crisis event which consists of all the different of type of crisis like elopee, jumper, and much more.
+From the years 2014 to 2023, Toronto has the highest number of PIC calls for the Person In crisis events which consists of all the different types of crises like elopee, jumper, and much more.
 The second highest category is suicide related which includes both suicide and suicide attempted and is a total of 91501 calls.
-The last category is Overdose.The total number of calls recorded in this dataset is 291991.
-The results can be seen @fig-events.
+The last category is Overdose.
+The total number of calls recorded in this dataset is 291991.
+The results can be seen at @fig-events.
 
@@
 @fig-hood shows the relationship between neighborhood and suicide PIC calls.
-There are around 174 neighborhoods divided in Hood_158 system and the graph shows around 158 neighborhoods have reported a PIC call for suicide-related events.
-The most number of calls were received from Hood 170 which were about 1022 suicide related calls and second most calls from received from the area Hood_73 which were around 858 calls.
-The Hood_170 is Yonge-Bay
-Corridor area and Hood_73 is Moss Park area in Old City of Toronto[@resource2].
+There are around 174 neighborhoods divided in the Hood_158 system and the graph shows around 158 neighborhoods have reported a PIC call for suicide-related events.
+The most number of calls were received from Hood 170 which were about 1022 suicide-related calls and the second most calls from received from the area Hood_73 which were around 858 calls.
+The Hood_170 is Yonge-Bay Corridor area and Hood_73 is Moss Park area in the Old City of Toronto[@resource2].
 
@@
                 "Thursday" = "orange", "Friday" = "brown", "Saturday" = "pink")
+
 # Create ggplot objects for days and time
+# code inspiration from: https://github.com/alyssaschleifer/ttc-subway-delay-analysis/blob/main/outputs/paper/Paper.Rmd
 
@@
 
-Upon thorough examination of the dataset, several significant findings emerge.
-Overall, suicide PIC calls show a very important and cruicial relationship with the other factors such as time of the day, day of the week, and geographical area.
+Upon thorough examination of the data set, several significant findings emerge.
+Overall, suicide PIC calls show a very important and crucial relationship with the other factors such as time of the day, day of the week, and geographical area.
 As highlighted in the Canadian Journal of Psychiatry[@resource3], suicide remains a noteworthy cause of death in Toronto, necessitating a nuanced understanding of the factors contributing to suicide and the development of targeted prevention strategies.
@@
 
-# Convert the table to a data frame
-#df <- as.data.frame.matrix(table_data)
-
-# Reorder days of the week
-#df$Day <- factor(row.names(df), levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
-
-# Melt the data for ggplot
-#melted_df <- melt(df, id.vars = "Day")
-
-# Create a heatmap
-#ggplot(melted_df, aes(x = variable, y = Day, fill = value)) +
- # geom_tile() +
-#  scale_fill_gradient(low = "white", high = "darkred") +
- # labs(title = "Suicide Calls by Day of the Week and Hour of the Day",
-#       x = "Hour of the Day",
- #      y = "Day of the Week") +
- # theme_minimal()
-
 ```
@@
 
-However, it is crucial to acknowledge and discuss the inherent biases and limitations within the dataset.
+However, it is crucial to acknowledge and discuss the inherent biases and limitations within the data set.
 The presence of numerous "NAS" (Not A Statement) assigned rows in the raw data introduces the possibility of errors in the graphs and analyses, potentially diverting the accuracy of the findings.
-Additionally, the dataset lacks explicit information regarding the reasons for caller apprehension, hindering a direct correlation between apprehension and the nature of events.
+Additionally, the data set lacks explicit information regarding the reasons for caller apprehension, hindering a direct correlation between apprehension and the nature of events.
 

### 2024-01-25 outputs/paper/paper.qmd
final changes

@@
 library(ggrepel)
+library(reshape2)
 
@@
 
-Toronto's Person In Crisis lines are an invaluable community asset that provides free, anonymous assistance to people experiencing a serious crisis.
+Toronto's Person In Crisis(PIC) lines are an invaluable community asset that provides free, anonymous assistance to people experiencing a serious crisis.
 The frequency of crises in today’s cities, particularly those involving populations at risk (PICs), highlights the need to better understand the dynamics behind such incidents.
@@
 A variety of event types, including attempt suicide, the person in crisis, elopee, jumper, overdose, and threatened suicide, are included in the data set.
-The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014,2015.
+The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014 and 2015.
 In this paper, all the neighborhood (Hood_158) data is used as an estimand to get a better understanding of all the hotspots of the PIC calls in the Toronto neighborhood.
@@
 # graph displays frequency of all the different type of events
+# More appealing to show through a graph rather than a table, numbers are easier to visualize and put into perspective
  ggplot(cleaned_data, aes(x = event_type)) +
@@
   theme_minimal() +
-  theme(legend.position = "top",        # Change legend position
+  theme(legend.position = "none",        # Change legend position
         legend.title = element_blank()) +
@@
 
+# A heatmap for the data visualization of days and time of PIC suicide calls
+
+# code inspiration from: https://r-graph-gallery.com/215-the-heatmap-function.html
+
+table_data <- table(cleaned_data_suicide1$event_dow, cleaned_data_suicide1$event_hour)
+df <- as.data.frame.matrix(table_data)
+ # Reorder days of the week
+df$Day <- factor(row.names(df), levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+# Melt the data for ggplot
+melted_df <- melt(df, id.vars = "Day")
+# Create a heatmap
+ggplot(melted_df, aes(x = variable, y = Day, fill = value)) +
+  geom_tile() +
+  scale_fill_gradient(low = "white", high = "darkred") +
+  labs(title = "Suicide Calls by Day of the Week and Hour of the Day",
+       x = "Hour of the Day",
+      y = "Day of the Week") +
+  theme_minimal() + 
+  theme(axis.text.x=element_text(size=6))
+
 # Create ggplot objects for days and time
@@
 
-time <- cleaned_data %>%
-  mutate(Time=as.POSIXct(event_hour, format="%H")) %>%
-  ggplot(aes(Time, fill=fct_infreq(event_dow))) +
-  geom_density(adjust=0.5, position="stack") +
-  labs(x = "Hour", y = "Density", fill= "Day of Week") +
-  scale_x_continuous(breaks = seq(0, 24, by = 2)) +
-  scale_fill_manual(values=day_colors) +
-  theme(axis.text.x=element_text(size=8)) +
-  ggtitle("Frequency of Suicide Calls")
+# This is another example of the map for visualization, run it if you want a better understanding with colors.
 
-time
+#time <- cleaned_data %>%
+#  mutate(Time=as.POSIXct(event_hour, format="%H")) %>%
+#  ggplot(aes(Time, fill=fct_infreq(event_dow))) +
+#  geom_density(adjust=0.5, position="stack") +
+#  labs(x = "Hour", y = "Density", fill= "Day of Week") +
+#  scale_x_continuous(breaks = seq(0, 24, by = 5)) +
+#  scale_fill_manual(values=day_colors) +
+#  theme(axis.text.x=element_text(size=8)) +
+#  ggtitle("Frequency of Suicide Calls")
+
+#time
 
@@
   summarise(suicide_frequency = n()) %>% # to get the frequency of the suicide calls
-  ggplot(aes(x = factor(event_hour), y = suicide_frequency, label = "Calls")) + geom_point() + 
-  ggrepel::geom_text_repel() + 
+  ggplot(aes(x = factor(event_hour), y = suicide_frequency)) + geom_point() + 
 labs(
@@
 Thursday is the day of the week that receives most of the calls with Sunday being the day that receives the least calls.
-The hours between 1300 to 1500 receive a high amount of calls compared to the other time of the day which shows a direct relation between the day and time and suicide related calls.
+The hours between 1500 to 1600 receive a high amount of calls compared to the other time of the day which shows a direct relation between the day and time and suicide related calls.
 

### 2024-01-25 scripts/00-simulate_data.R

@@ -9,6 +9,7 @@
 #### Workspace setup ####
 library(tidyverse)
 library(opendatatoronto)
+library(ggplot2)
 
 #### Simulate data ####
 set.seed(398)
@@ -17,7 +18,9 @@ set.seed(398)
 # was made or not and the Toronto Hood 158 number the call is associated with
 simulated_data <-tibble(
   "Event_id" = 1:1000,
-  "Event_type" = sample(c("Suicide-related", "Person in Crisis", "Elopee", "Jumper", "Overdose", "Threaten Suicide"), 1000, replace = TRUE),
+  "Event_type" = sample(c("Suicide-related", "Person in Crisis", "Elopee", 
+                          "Jumper", "Overdose", "Threaten Suicide"), 1000, 
+                        replace = TRUE),
   "Apprehension_made" = sample(c("Yes", "No"), 1000, replace = TRUE),
   "Hood_158" = sample(100:200, 1000, replace = TRUE),
 )
@@ -40,12 +43,14 @@ event_freq
 
 ### Visualize data ###
 
-#based on code from: https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html #Explore
+#based on code from: 
+# https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html #Explore
 
 # Visualize event types and their frequencies
 ggplot(simulated_data, aes(x = Event_type)) +
   geom_bar(fill = "pink") +
-  labs(title = "Event Types and Their Frequencies", x = "Event Type", y = "Frequency", color = "Country") +
+  labs(title = "Event Types and Their Frequencies", x = "Event Type", 
+       y = "Frequency", color = "Country") +
 scale_color_brewer(palette = "Set1") +
   theme(legend.position = "bottom")
 
@@ -53,20 +58,24 @@ scale_color_brewer(palette = "Set1") +
 # Visualize event types based on neighborhoods
 #ggplot(simulated_data, aes(x = Event_type, fill = NEIGHBOURHOOD)) +
 #  geom_bar(position = "dodge") +
-#  labs(title = "Event Types Based on Neighborhoods", x = "Event Type", y = "Frequency", fill = "Neighborhood")
+#  labs(title = "Event Types Based on Neighborhoods", x = "Event Type",
+# y = "Frequency", fill = "Neighborhood")
 
 # Visualize event type based on Hood_158 while using the stack formation
 ggplot(simulated_data, aes(x = Event_type, fill = factor(Hood_158))) +
[…]
